I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-19 Paper/Paper_noProof.tex
Section 3

MILP instances can be linearly relaxed into LP over-abstraction, where variables which were integers in 
${0,1}$ are relaxed to be real in interval $[0,1]$, using otherwise the same encoding. 
Solving LP instances being Polynomial time, much more efficient solution could be found, but at the price of a lack of accuracy (e.g. with not as tight bounds). This is called the {\em LP abstraction}.

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate that the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(n) \geq 0$ and $ReLU(n) \geq n$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
 Given $n \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
 1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
 \vspace{-0.3cm}
 \begin{align*}
	 & 1) \quad \hat{n} \geq n \quad \wedge \quad \hat{n} \geq 0, \quad \wedge \quad \hat{n} \leq \beta \cdot a \quad \wedge \quad \hat{n} \leq n-\alpha \cdot (1-a), \, a \in [0,1] \\
	 \text{and} \quad  & 2)  \quad \hat{n} \geq n \quad \wedge \quad \hat{n} \geq 0 \quad \wedge \quad \hat{n} \leq \beta \frac{n-\alpha}{\beta-\alpha}
\end{align*} 
\end{proposition}

@@

It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the variable $a \in  [0,1]$, with $n \in [\alpha,\beta]$ fixed. We have $\hat{n}$ in 1) upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, n - \alpha (1-a)))$, and this bound can be reached. 
The function $min(\beta \cdot a, n - \alpha (1-a))$ reaches its maximum for 
$\beta \cdot a = n - \alpha (1-a)$, that is 
$(\beta - \alpha) a = n - \alpha$, i.e. 
$$ a = \frac{n - \alpha}{\beta-\alpha}$$

It thus gives an upper bound at 
$\beta \cdot a = \beta \frac{n - \alpha}{\beta-\alpha}$, which is exactly the function used in DeepPoly.
\hfill $\square$
\end{proof}

@@

In this section, we explore the key factor for the loss of accuracy in value abstraction methodologies, as those we presented before (except for full MILP which is exact but too costly). 
Namely, we uncover that \emph{compensating pair of paths} are the reason for the loss of precision.

Formally, a pair $(\pi,\pi')$ of paths is {\em compensating} if:
\begin{itemize}
 \item they have the same starting node $a$ (called {\em source} state) and ending node $b$ (called {\em target} state of $(\pi,\pi')$),
 \item they are disjoints (the only common nodes are the source and the target),
 \item the weights satisfy $weight(\pi') < 0 < weight(\pi)$.
\end{itemize}

For instance, on Fig.~\ref{fig1}, the paths $\pi= x_2 x_3 x_5$ has weight $1$, while the
path $\pi'= x_2 x_4 x_5$ has weight $-2$, hence $(\pi,\pi')$ is a compensating pair of paths.
We call a path $\pi$ {\em compensating} if there exists a path $\pi'$ such that either $(\pi,\pi')$ or 
$(\pi',\pi)$ is compensating.

Compensating pair of paths will make precision of the target state reduce when using overabstraction of values. Here, $\Val(x_5) \leq 4$ (the value is reached for $x_1=1$ and $x_2=-1$). However, Box computes an upper bound of $6$, and while DeepPoly recovers some of the dependency in $x_2$, it can only cancel out some of $x_2$ value due to the ReLU over-abstraction (contribution of $\frac{-x_2+4}{2} \leq 2.5$ instead of exact $-x_2 \leq 1$), obtaining a bound $\Val(x_5) \leq 5$ instead of $4$.

Intuitively, the target state receives the weight from the source state from the different paths from the source to the target. As the weights of the paths is of opposite sign, some of these weights will cancel out, by an amount up to $min(|weight(\pi')|,|weight(\pi')|)$. Further, no only do they cancel $min(|weight(\pi')|,|weight(\pi')|)$ out, but they also do not accumulate, in the same way as $0= |1+(-1)| \leq |1|+|-1|=2$ from an inaccurate analysis. The amount of compensation can however be reduced by the ReLU functions, because of clipping by the ReLU function. For instance, consider that the inputs $x_1,x_2$ lies in $[-1,-0]$. Then the weight at $\hat{x}_3$ will be always 0, no matter the input, and there will be no compensation to $weight(\pi')$.

@@

Hence, compensation is {\em one} factor of inaccuracy. We prove now that it is actually the {\em only} reason for inaccuracy, in the following meaning.
If there is no compensating path in the DNN, then any method/approximation at least as accurate as the {\em Box} abstraction, e.g. $\overline{\text{DeepPoly}}$ or {\em LP}, computes the exact upper and lower bounds for every node, that is:

@@

\begin{theorem}
	\label{th1}
	Let $n$ be any node of a DNN with {\em no} compensating path. For $[\alpha,\beta]$ the bounds computed by the {\em Box} abstraction for $n$, we have that for all $\gamma \in [\alpha,\beta]$, there exists an input $\vx$ such that $\val_{\vx}(n)=\gamma$.
\end{theorem}

The sketch of proof can be found in Section \ref{sec.proofs}. 
Notice that standard {\em DeepPoly} does not satisfies the following theorem (it does not necessarily refine {\em Box}), and therefore we are not sure PRIMA or $\beta$-CROWN who refine DeepPoly do, though it would be easy to fix by using $\overline{\text{DeepPoly}}$. Anyway, in practice, it is very unlikely to have a network without any compensating path, it is more of a theoretical results show in that compensation is the reason for inaccuracy.

>>>>

MILP instances can be linearly relaxed into LP over-abstraction, where variables which were integers in $\{0,1\}$ (binary) are relaxed to be real in interval $[0,1]$, using otherwise the same encoding. Solving LP instances being Polynomial time, optimizing it is much more efficient, but at the price of a lack of accuracy (e.g. with not as tight bounds). This is called the {\em LP abstraction}.

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate than the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(x) \geq 0$ and $ReLU(x) \geq x$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
 Given $x \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
 1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
 \vspace{-0.3cm}
 \begin{align*}
	 & 1) \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a), \, a \in [0,1] \\
	 \text{and} \quad  & 2)  \quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0 \quad \wedge \quad \hat{x} \leq \beta \frac{x-\alpha}{\beta-\alpha}
\end{align*} 
\end{proposition}

@@

It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the linear variable $a \in  [0,1]$, with $x \in [\alpha,\beta]$ fixed. We have $\hat{x}$ in 1) is uppered bounded by $max_{a \in [0,1]} (min(\beta \cdot a, x - \alpha (1-a)))$, and this bound can be reached. 
The function $min(\beta \cdot a, x - \alpha (1-a))$ reaches its maximum for 
$\beta \cdot a = x - \alpha (1-a)$, that is 
$(\beta - \alpha) a = x - \alpha$, i.e. 
$$ a = \frac{x - \alpha}{\beta-\alpha}$$

It thus gives an upper bound at 
$\beta \cdot a = \beta \frac{x - \alpha}{\beta-\alpha}$, which is exactly the function used in DeepPoly.
\hfill $\square$
\end{proof}

@@

In this section, we explore the key factor for the loss of accuracy in value abstraction methodologies, as those we presented before (except for full MILP which is exact but 
not efficient). 
Namely, we uncover that \emph{compensating pair of paths} are the reason for the loss of precision. A pair $(\pi,\pi')$ of paths is {\em compensating} if:
\begin{itemize}
 \item they have the same starting node $a$ (called {\em source} state) and ending node $b$ (called {\em target} state of $(\pi,\pi')$),
 \item they are disjoints (the only common nodes are the source and the target),
 \item the weights satisfy $weight(\pi') < 0 < weight(\pi)$.
\end{itemize}

For instance, on Fig.~\ref{fig1}, the paths $\pi= n_2 n_3 n_5$ has weight $1$, while the
path $\pi'= n_2 n_4 n_5$ has weight $-2$, hence $(\pi,\pi')$ is a compensating pair of paths.
We call a path $\pi$ {\em compensating} if there exists a path $\pi'$ such that either $(\pi,\pi')$ or $(\pi',\pi)$ is compensating.

Compensating pair of paths will make precision of the target state reduce when using overabstraction of values. Here, $x_5 \leq 4$, which is reached for $x_1=1$ and $x_2=-1$: 
$\val_{(1,-1)}(n_5)=4$. However, Box computes an upper bound of $6$, and while DeepPoly recovers some of the dependency in $x_2$, it can only cancel out some of $x_2$ value due to the ReLU over-abstraction (contribution of $\frac{-x_2+3}{2} \leq 2$ instead of exact $-x_2 \leq 1$), obtaining $x_5 \leq 5$ instead of the precise $x_5 \leq 4$.

Intuitively, the target state receives the weight from the source state from the different paths from the source to the target. As the weights of the paths is of opposite sign, some of these weights will cancel out, by an amount up to $min(|weight(\pi')|,|weight(\pi')|)$. 
Further, that also avoid accumulating the error, in the same way as $0= |1+(-1)| \leq |1|+|-1|=2$ from an inaccurate analysis. The amount of compensation can however be reduced by the ReLU functions, because of clipping by the ReLU function. For instance, consider that the inputs $x_1,x_2$ lies in $[-1,-0]$. Then $\hat{x}_3$ will always be 0 for any input $\vx=(x_1,x_2) \in [-1,-0] \times [-1,-0]$, and $weight(\pi')$ will not compensate $weight(\pi')$.

@@

Hence, compensation is {\em one} factor of inaccuracy. We prove now that it is actually the {\em only} reason for inaccuracy, in the following meaning.
If there is no compensating path in the DNN, then any method/approximation at least as accurate as the Box abstraction, e.g. $\overline{\text{DeepPoly}}$ or LP, computes the exact upper and lower bounds for every node, that is:

@@

\begin{theorem}
	\label{th1}
	Let $n$ be any node of a DNN with {\em no} compensating path. For $[\alpha,\beta]$ the bounds computed by the {\em Box abstraction} for $n$, we have that for all $\gamma \in [\alpha,\beta]$, there exists an input $\vx$ such that $\val_{\vx}(n)=\gamma$.
\end{theorem}

The sketch of proof can be found in Section \ref{sec.proofs}. 
Notice that standard DeepPoly does not satisfies Theorem \ref{th1} (it does not necessarily refine Box), and therefore we are not sure PRIMA or $\beta$-CROWN that refine DeepPoly satisfies Theorem \ref{th1}, though it would be easy to fix by using $\overline{\text{DeepPoly}}$. Anyway, in practice, it is very unlikely to have a network without any compensating path, it is more of a theoretical results show in that compensation is the reason for inaccuracy.
